\bf Estimating the number of clusters using cross-validation

### 2017-02-08 cvclust.tex
change line endings (DOS to UNIX)
add more clarification on the assumptions

It will suffice to prove self-consistency for a single fold of the
cross-validation procedure.  As in section~\ref{sec:gabriel-cv-algorithm}
we assume that the $P$ variables of the data set have been partitioned into
$p$ predictor variables represented in vector~$X$ and $q$ response variables
represented in vector~$Y$.  The $N$ observations have been divided into two
sets: $n$ train observations and $m$ test observations.  The following theorem
gives conditions for Gabriel cross-validation to recover the true number of
clusters in the absence of noise.

@@

Given $\mu(g) = \bigl(\muX(g),\muY(g)\bigr)$ are distinct for $g = 1, \dotsc, K$,
~\ref{asn:self-consistency-distinct-mux} and ~\ref{asn:self-consistency-distinct-muy}
are reasonable given not too small dimension $p$ and $q$.
The result statement of Proposition~\ref{prop:self-consistency} follows from 
Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
will now state and prove.

@@

@@

We will now analyze a simple two-cluster setting, and derive conditions for
Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$. In the
statement of the result, $\Phi(\cdot)$ and $\varphi(\cdot)$ denote the
standard normal cumulative distribution function and density, respectively.

\begin{proposition}\label{prop:twoclust}
Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
cross-validation, where each $(X,Y)$ pair in $\R^2$ is an independent draw
from an equiprobable mixture of two multivariate normal distributions with
identity covariance. Suppose that the first mixture component has mean
$\mu = (\muX, \muY)$ and the second has mean $-\mu = (-\muX, -\muY)$,
where $\muX \geq 0$ and $\muY \geq 0$. 
If 
\(
  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX),
\)
then $\CV(2) < \CV(1)$ with probability tending to one as $m$ and $n$ increase.
\end{proposition}

@@

When $k=1$, the $k$-means centroid is equal to the sample mean $\bar Y_n = (1/n)
\sum_{i=1}^{n} Y_i$, approximately equal to 
$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error is
\[
  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
         = 1 + (\mu^Y)^2 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
\]
Thus, if $a + 2 \muY - 4 \muY \Phi(\muX) < 0$, then $\CV(2) < \CV(1)$ with
probability tending to one as $m$ and $n$ increase. Substituting the
expression for $\E(Y \mid Y > 0)$ in place of $a$, the inequality holds
precisely when
\(
  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX).
\)
\end{proof}
The inequality in Proposition~\ref{prop:twoclust} means that the proposed method
would correctly pick $k = 2$ given that $(\muX, \muY)$ and 
$(-\muX, -\muY)$ are not too close. 
We confirm the result of Proposition~\ref{prop:twoclust} with a simulation.
We perform $10$ replicates for each $(\mu^X, \mu^Y)$ pair, sweeping over
a two-dimensional grid of values in the domain $[0,3] \times [0,3]$, with step
size $0.1$ in each dimension.  In each
replicate, we generate $N=20000$ observations from an equiprobably mixture
of multivariate normal distributions with identity covariance, with one
component having mean $(\muX, \muY)$ and
the other component having mean $(-\muX, -\muY)$. We perform a single $2 \times 2$
fold of Gabriel cross-validation and report the number of times
(out of $10$ replicates) where $k=2$ is selected by the algorithm instead of $k=1$.
Figure~\ref{fig:overlap-color_plot} shows the frequency with which $k=2$ is selected by
the algorithm for each $(\mu^X, \mu^Y)$ pair. Darker (red) colors indicate
higher numbers (close to $10$), situations where $k = 2$ is selected more
often than $k = 1$. Ligher (blue) colors indicate that $k = 1$ is preferred.
We can see the simulation result perfectly align with the
theoretical curve (the black line), which separates the $k=2$ zone from the
$k=1$ zone.

>>>>

It will suffice to prove self-consistency for a single fold of the
cross-validation procedure.  As in section~\ref{sec:gabriel-cv-algorithm} we
assume that the $P$ variables of the data set have been partitioned into $p$
predictor variables represented in vector~$X$ and $q$ response variables
represented in vector~$Y$.  The $N$ observations have been divided into two
sets: $n$ train observations and $m$ test observations.  We state the
assumptions for the self-consistency result in terms of a specific split; for
the result to hold in general, with high probability, these assumptions would
have to hold with high probability for a random split. The following theorem
gives conditions for Gabriel cross-validation to recover the true number of
clusters in the absence of noise.

@@

The proposition states that our method works well in the absence of noise,
when each observation is equal to its cluster center.  The essential
assumption here is assumption~\ref{asn:self-consistency-noiseless}, which states
that there is no noise.  If we are willing to assume, say, that the cluster
centers~$\mu(g) = \bigl(\muX(g),\muY(g)\bigr)$ for $g = 1, \dotsc, K$ were
randomly drawn from a distribution with a density over $\R^{p+q}$, then
assumptions~\ref{asn:self-consistency-distinct-mux}
and~\ref{asn:self-consistency-distinct-muy} will hold with probability one for
all splits of the data. Likewise, if the clusters are not too small (relative
to $n$ and $m$), then assumptions~\ref{asn:self-consistency-train}
and~\ref{asn:self-consistency-test} will likely hold for a random split of the
data into test and train.

Proposition~\ref{prop:self-consistency} follows from
Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
now state and prove.

@@

Proposition~\ref{prop:single-2d} tells us that Gabriel cross-validation fails
when there is strong correlation between the variables. To get around this, in
practice we will transform the data to reduce correlation before performing
cross-validation. We detail this approach in Section~\ref{sec:corr-correct}.

@@

We will now analyze a simple two-cluster setting, and derive conditions for
Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$.  The
main assumption of the proposition is that the cluster centers are not too
close. The precise definition of ``too close'' is stated in terms of
$\Phi(\cdot)$ and $\varphi(\cdot)$, the standard normal cumulative
distribution function and density, respectively.  The inequality is hard to
interpret directly, but we show the boundary between ``too close'' and ``well
separated'' in Fig.~\ref{fig:overlap-color_plot}, after the proof of the
proposition.

\begin{proposition}\label{prop:twoclust}
Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
cross-validation, where each $(X,Y)$ pair in $\R^2$ is an independent draw
from an equiprobable mixture of two multivariate normal distributions with
identity covariance. Suppose that the first mixture component has mean
$\mu = (\muX, \muY)$ and the second has mean $-\mu = (-\muX, -\muY)$,
where $\muX \geq 0$ and $\muY \geq 0$. 
If the cluster centers are well separated, specifically such that
\(
  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX),
\)
then $\CV(2) < \CV(1)$ with probability tending to one as $m$ and $n$ increase.
\end{proposition}

@@

When $k=1$, the $k$-means centroid is equal to the sample mean $\bar Y_n = (1/n)
\sum_{i=1}^{n} Y_i$, approximately equal to 
$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error is
\[
  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
         = 1 + (\mu^Y)^2 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
\]
Thus, if $a + 2 \muY - 4 \muY \Phi(\muX) < 0$, then $\CV(2) < \CV(1)$ with
probability tending to one as $m$ and $n$ increase. Substituting the
expression for $\E(Y \mid Y > 0)$ in place of $a$, the inequality holds
precisely when
\(
  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX).
\)
\end{proof}
We confirm the result of Proposition~\ref{prop:twoclust} with a simulation.
We perform $10$ replicates for each $(\mu^X, \mu^Y)$ pair, sweeping over
a two-dimensional grid of values in the domain $[0,3] \times [0,3]$, with step
size $0.1$ in each dimension.  In each
replicate, we generate $N=20000$ observations from an equiprobably mixture
of multivariate normal distributions with identity covariance, with one
component having mean $(\muX, \muY)$ and
the other component having mean $(-\muX, -\muY)$. We perform a single $2 \times 2$
fold of Gabriel cross-validation and report the number of times
(out of $10$ replicates) where $k=2$ is selected by the algorithm instead of $k=1$.
Figure~\ref{fig:overlap-color_plot} shows the frequency with which $k=2$ is selected by
the algorithm for each $(\mu^X, \mu^Y)$ pair. Darker (red) colors indicate
higher numbers (close to $10$), situations where $k = 2$ is selected more
often than $k = 1$. Ligher (blue) colors indicate that $k = 1$ is preferred.
We can see the simulation result perfectly align with the
theoretical curve (the black line), which separates the $k=2$ zone from the
$k=1$ zone.
